---
title: "Predicting Developer Job Satisfaction: A Multi-Model Classification Study"
author: "Group W11G03"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: hide
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 8, fig.height = 5, fig.align = "center"
)
library(caret)
library(gbm)
library(tidyverse)
library(ggplot2)
library(knitr)
library(kableExtra)
```


# Data Preprocessing {.tabset}

## Overview

This study uses the **Stack Overflow 2024 Developer Survey** (53,921 responses, 172 variables) to predict developer job satisfaction. The original 0–10 numeric `JobSat` score is discretised into a **3-class target variable**:

| Class | JobSat Range | Interpretation |
|:------|:------------|:---------------|
| Dissatisfied | 0 – 3 | Very Dissatisfied + Dissatisfied |
| Neutral | 4 – 6 | Neutral |
| Satisfied | 7 – 10 | Satisfied + Very Satisfied |

The 3-class formulation was chosen over 5-class because adjacent 5-class boundaries differ by only 1–2 points on a subjective scale and are inherently ambiguous. Merging reduces the class imbalance ratio from approximately 30:1 to approximately 11:1 and produces clearer decision boundaries for all models.

## Preprocessing Code and Summary

```{r preprocessing, eval=TRUE}
# ============================================================
# Shared preprocessing pipeline
# All parameters learned from training set only (no data leakage)
# ============================================================

set.seed(123)
data_path <- "survey_results_public.csv"
df <- read_csv(data_path, show_col_types = FALSE)

# --- Helper functions ---

convert_years <- function(x) {
  x <- as.character(x)
  x[x == "Less than 1 year"] <- "0.5"
  x[x == "More than 50 years"] <- "51"
  as.numeric(x)
}

safe_median <- function(x) {
  med <- median(x, na.rm = TRUE)
  ifelse(is.na(med), 0, med)
}

safe_quantile <- function(x, prob = 0.99) {
  if (all(is.na(x))) return(0)
  q <- quantile(x, probs = prob, na.rm = TRUE)
  ifelse(is.na(q), max(x, na.rm = TRUE), q)
}

safe_remove_vars <- function(data, vars) {
  vars <- vars[!is.na(vars) & vars != "" & vars %in% names(data)]
  if (length(vars) > 0) data <- data %>% select(-any_of(vars))
  data
}

is_near_zero_var <- function(x, freq_cut = 95 / 5, unique_cut = 10) {
  x <- as.character(x); x <- x[!is.na(x)]
  if (length(x) == 0) return(TRUE)
  tab <- sort(table(x), decreasing = TRUE)
  if (length(tab) == 1) return(TRUE)
  freq_ratio <- as.numeric(tab[1] / tab[2])
  percent_unique <- length(unique(x)) / length(x) * 100
  freq_ratio > freq_cut && percent_unique < unique_cut
}

# --- Step 1: Select 22 candidate predictors + create 5-class target ---

gbm_raw <- df %>%
  mutate(
    JobSat = as.numeric(JobSat),
    JobSat_Class = case_when(
      JobSat <= 1  ~ "Very_Dissatisfied",
      JobSat <= 3  ~ "Dissatisfied",
      JobSat <= 6  ~ "Neutral",
      JobSat <= 8  ~ "Satisfied",
      JobSat <= 10 ~ "Very_Satisfied",
      TRUE ~ NA_character_
    ),
    JobSat_Class = factor(JobSat_Class,
      levels = c("Very_Dissatisfied","Dissatisfied","Neutral",
                 "Satisfied","Very_Satisfied")),
    YearsCode = convert_years(YearsCode),
    WorkExp = as.numeric(WorkExp),
    ConvertedCompYearly = as.numeric(ConvertedCompYearly)
  ) %>%
  select(any_of(c(
    "JobSat_Class",
    "Age","EdLevel","Employment","RemoteWork","WorkExp","YearsCode",
    "OrgSize","ICorPM","Country","DevType","Industry",
    "AISelect","AISent","AIAcc","AIThreat",
    "NewRole","TechEndorseIntro","PurchaseInfluence",
    "ToolCountWork","ToolCountPersonal","ConvertedCompYearly"
  ))) %>%
  filter(!is.na(JobSat_Class))

# --- Step 2: Stratified 80/20 train-test split BEFORE any preprocessing ---

train_index <- createDataPartition(gbm_raw$JobSat_Class, p = 0.8, list = FALSE)
train_raw <- gbm_raw[train_index, ]
test_raw  <- gbm_raw[-train_index, ]

# --- Step 3: Fit preprocessing rules on training data only ---

fit_gbm_preprocess <- function(train_data) {
  missing_rates <- sapply(train_data, function(x) mean(is.na(x)))
  high_missing_vars <- names(missing_rates)[
    missing_rates > 0.80 & names(missing_rates) != "JobSat_Class"]
  high_missing_vars <- high_missing_vars[!is.na(high_missing_vars) & high_missing_vars != ""]
  train_data_reduced <- safe_remove_vars(train_data, high_missing_vars)

  numeric_vars <- intersect(c("WorkExp","YearsCode","ToolCountWork","ToolCountPersonal"),
                            names(train_data_reduced))
  numeric_medians <- list()
  for (v in numeric_vars) numeric_medians[[v]] <- safe_median(train_data_reduced[[v]])

  workexp_cap <- if ("WorkExp" %in% names(train_data_reduced))
    safe_quantile(train_data_reduced$WorkExp, 0.99) else NA
  yearscode_cap <- if ("YearsCode" %in% names(train_data_reduced))
    safe_quantile(train_data_reduced$YearsCode, 0.99) else NA
  salary_cap <- if ("ConvertedCompYearly" %in% names(train_data_reduced))
    safe_quantile(train_data_reduced$ConvertedCompYearly, 0.99) else NA

  if ("ConvertedCompYearly" %in% names(train_data_reduced)) {
    salary_temp <- train_data_reduced$ConvertedCompYearly
    salary_temp[salary_temp < 0] <- NA
    salary_temp <- ifelse(salary_temp > salary_cap, salary_cap, salary_temp)
    log_salary_median <- safe_median(log1p(salary_temp))
  } else { log_salary_median <- 0 }

  multi_response_candidates <- intersect(
    c("DevType","Employment","AISelect","AISent","AIAcc","AIThreat",
      "NewRole","TechEndorseIntro","PurchaseInfluence"),
    names(train_data_reduced))
  multi_response_vars <- c(); multi_response_terms <- list()
  for (v in multi_response_candidates) {
    x <- as.character(train_data_reduced[[v]])
    if (any(str_detect(x[!is.na(x)], fixed(";")))) {
      multi_response_vars <- c(multi_response_vars, v)
      sv <- trimws(unlist(strsplit(x[!is.na(x)], ";", fixed = TRUE)))
      sv <- sv[sv != ""]
      vc <- sort(table(sv), decreasing = TRUE)
      multi_response_terms[[v]] <- names(vc)[1:min(15, length(vc))]
    }
  }

  categorical_vars <- setdiff(
    intersect(c("Age","EdLevel","RemoteWork","OrgSize","ICorPM","Country","Industry",
                "AISelect","AISent","AIAcc","AIThreat","Employment","DevType",
                "NewRole","TechEndorseIntro","PurchaseInfluence"),
              names(train_data_reduced)),
    multi_response_vars)
  top_levels <- list()
  for (v in categorical_vars) {
    x <- as.character(train_data_reduced[[v]]); x[is.na(x)] <- "Missing"
    lc <- sort(table(x), decreasing = TRUE)
    top_levels[[v]] <- names(lc)[1:min(15, length(lc))]
  }

  list(high_missing_vars = high_missing_vars,
       numeric_vars = numeric_vars, numeric_medians = numeric_medians,
       workexp_cap = workexp_cap, yearscode_cap = yearscode_cap,
       salary_cap = salary_cap, log_salary_median = log_salary_median,
       categorical_vars = categorical_vars, top_levels = top_levels,
       multi_response_vars = multi_response_vars, multi_response_terms = multi_response_terms)
}

# --- Step 4: Apply preprocessing to both train and test ---

apply_gbm_preprocess <- function(data, prep) {
  data <- safe_remove_vars(data, prep$high_missing_vars)
  if ("WorkExp"             %in% names(data)) data$Missing_WorkExp             <- ifelse(is.na(data$WorkExp), 1, 0)
  if ("YearsCode"           %in% names(data)) data$Missing_YearsCode           <- ifelse(is.na(data$YearsCode), 1, 0)
  if ("ConvertedCompYearly" %in% names(data)) data$Missing_ConvertedCompYearly <- ifelse(is.na(data$ConvertedCompYearly), 1, 0)
  if ("OrgSize"             %in% names(data)) data$Missing_OrgSize             <- ifelse(is.na(data$OrgSize), 1, 0)
  if ("WorkExp" %in% names(data)) {
    data$WorkExp[data$WorkExp < 0] <- NA
    data$WorkExp <- ifelse(!is.na(data$WorkExp) & data$WorkExp > prep$workexp_cap, prep$workexp_cap, data$WorkExp)
    data$WorkExp[is.na(data$WorkExp)] <- prep$numeric_medians[["WorkExp"]]
  }
  if ("YearsCode" %in% names(data)) {
    data$YearsCode[data$YearsCode < 0] <- NA
    data$YearsCode <- ifelse(!is.na(data$YearsCode) & data$YearsCode > prep$yearscode_cap, prep$yearscode_cap, data$YearsCode)
    data$YearsCode[is.na(data$YearsCode)] <- prep$numeric_medians[["YearsCode"]]
  }
  for (v in setdiff(prep$numeric_vars, c("WorkExp","YearsCode"))) {
    if (v %in% names(data)) data[[v]][is.na(data[[v]])] <- prep$numeric_medians[[v]]
  }
  if ("ConvertedCompYearly" %in% names(data)) {
    data$ConvertedCompYearly[data$ConvertedCompYearly < 0] <- NA
    data$ConvertedCompYearly <- ifelse(!is.na(data$ConvertedCompYearly) & data$ConvertedCompYearly > prep$salary_cap, prep$salary_cap, data$ConvertedCompYearly)
    data$LogConvertedCompYearly <- log1p(data$ConvertedCompYearly)
    data$LogConvertedCompYearly[is.na(data$LogConvertedCompYearly)] <- prep$log_salary_median
    data$ConvertedCompYearly <- NULL
  }
  for (v in prep$categorical_vars) {
    if (v %in% names(data)) {
      x <- as.character(data[[v]]); x[is.na(x)] <- "Missing"
      x[!(x %in% prep$top_levels[[v]])] <- "Other"
      data[[v]] <- factor(x, levels = unique(c(prep$top_levels[[v]], "Other", "Missing")))
    }
  }
  for (v in prep$multi_response_vars) {
    if (v %in% names(data)) {
      raw_x <- as.character(data[[v]]); raw_x[is.na(raw_x)] <- ""
      padded_x <- paste0(";", raw_x, ";")
      for (term in prep$multi_response_terms[[v]]) {
        data[[paste0(v, "_", make.names(term))]] <-
          ifelse(str_detect(padded_x, fixed(paste0(";", term, ";"))), 1, 0)
      }
      data[[v]] <- NULL
    }
  }
  data
}

gbm_prep   <- fit_gbm_preprocess(train_raw)
train_data <- apply_gbm_preprocess(train_raw, gbm_prep)
test_data  <- apply_gbm_preprocess(test_raw,  gbm_prep)

levels_order <- c("Very_Dissatisfied","Dissatisfied","Neutral","Satisfied","Very_Satisfied")
train_data$JobSat_Class <- factor(train_data$JobSat_Class, levels = levels_order)
test_data$JobSat_Class  <- factor(test_data$JobSat_Class,  levels = levels_order)

# --- Step 5: Feature engineering — LeavingScore ---

leaving_map <- c(
  "I have strongly considered changing my career and/or the industry I work in" = 4,
  "I have somewhat considered changing my career and/or the industry I work in" = 3,
  "I have transitioned into a new career and/or industry involuntarily"         = 2,
  "I have neither consider or transitioned into a new career or industry"       = 1,
  "I have transitioned into a new career and/or industry voluntarily"           = 0)

add_leaving_score <- function(data, median_value) {
  if ("NewRole" %in% names(data)) {
    nr_char <- as.character(data$NewRole)
    data$LeavingScore <- leaving_map[nr_char]
    data$LeavingScore[is.na(data$LeavingScore)] <- median_value
  }
  data
}

train_leaving_raw <- leaving_map[as.character(train_data$NewRole)]
leaving_median <- median(train_leaving_raw, na.rm = TRUE)
train_data <- add_leaving_score(train_data, leaving_median)
test_data  <- add_leaving_score(test_data,  leaving_median)

# --- Step 6: Feature selection ---

predictor_names <- setdiff(names(train_data), "JobSat_Class")
nzv_vars <- predictor_names[sapply(train_data[predictor_names], is_near_zero_var)]
nzv_vars <- nzv_vars[!is.na(nzv_vars) & nzv_vars != "" & nzv_vars %in% names(train_data)]
train_data <- safe_remove_vars(train_data, nzv_vars)
test_data  <- safe_remove_vars(test_data,  nzv_vars)

numeric_predictors <- names(train_data)[sapply(train_data, is.numeric) & names(train_data) != "JobSat_Class"]
numeric_predictors <- numeric_predictors[!is.na(numeric_predictors) & numeric_predictors != "" & numeric_predictors %in% names(train_data)]
numeric_predictors <- numeric_predictors[sapply(train_data[numeric_predictors], function(x) sd(x, na.rm = TRUE) > 0)]
high_corr_vars <- character(0)
if (length(numeric_predictors) > 1) {
  corr_matrix <- cor(train_data[numeric_predictors], use = "pairwise.complete.obs")
  corr_matrix[is.na(corr_matrix)] <- 0
  high_corr_index <- findCorrelation(corr_matrix, cutoff = 0.90, names = FALSE)
  if (length(high_corr_index) > 0) high_corr_vars <- numeric_predictors[high_corr_index]
}
high_corr_vars <- high_corr_vars[!is.na(high_corr_vars) & high_corr_vars != "" & high_corr_vars %in% names(train_data)]
train_data <- safe_remove_vars(train_data, high_corr_vars)
test_data  <- safe_remove_vars(test_data,  high_corr_vars)

# --- Step 7: Remap to 3-class ---

remap_to_3class <- function(x) {
  factor(case_when(
    x %in% c("Very_Dissatisfied","Dissatisfied") ~ "Dissatisfied",
    x == "Neutral" ~ "Neutral",
    x %in% c("Satisfied","Very_Satisfied") ~ "Satisfied",
    TRUE ~ NA_character_
  ), levels = c("Dissatisfied","Neutral","Satisfied"))
}

train_3 <- train_data; test_3 <- test_data
train_3$JobSat_Class <- remap_to_3class(train_data$JobSat_Class)
test_3$JobSat_Class  <- remap_to_3class(test_data$JobSat_Class)
```

The table below summarises each preprocessing step:

```{r preprocessing-table, eval=TRUE}
steps_df <- data.frame(
  Step = c("Initial selection", "High-missingness removal (>80% NA)",
           "Missing indicators", "Numeric capping + median imputation",
           "Salary log transformation", "Rare category merging (top 15 levels)",
           "Multi-response binarisation", "LeavingScore feature engineering",
           "Near-zero variance removal", "High-correlation removal (r > 0.90)"),
  Rationale = c(
    "Select 22 domain-relevant predictors from 172 survey columns",
    "Variables with excessive missingness cannot be reliably imputed",
    "Create binary flags for WorkExp, YearsCode, Salary, OrgSize missingness",
    "Cap at 99th percentile to limit outlier influence; impute with training median",
    "Right-skewed salary distribution normalised via log1p; NA filled with training median",
    "Prevent sparse levels from causing overfitting; rare categories grouped as 'Other'",
    "Semicolon-delimited columns (DevType, Employment, etc.) split into binary indicators",
    "Ordinal encoding of NewRole into 0-4 career-change propensity score",
    "Remove predictors where dominant class ratio exceeds 95/5",
    "Remove redundant numeric predictors to reduce multicollinearity"))
kable(steps_df, align = "ll") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = TRUE, font_size = 12)
```

After preprocessing: **`r nrow(train_3)`** training samples, **`r nrow(test_3)`** test samples, **`r ncol(train_3) - 1`** features.

```{r class-dist-plot, eval=TRUE, fig.height=3}
dist_df <- data.frame(
  Class = rep(c("Dissatisfied","Neutral","Satisfied"), 2),
  Set   = rep(c("Train","Test"), each = 3),
  Count = c(as.integer(table(train_3$JobSat_Class)), as.integer(table(test_3$JobSat_Class)))
)
ggplot(dist_df, aes(x = Class, y = Count, fill = Set)) +
  geom_col(position = "dodge", width = 0.6) +
  geom_text(aes(label = Count), position = position_dodge(0.6), vjust = -0.3, size = 3.5) +
  scale_fill_manual(values = c(Train = "#1565C0", Test = "#FB8C00")) +
  labs(title = "3-Class Target Distribution (Train vs Test)", x = NULL, y = "Count") +
  theme_minimal(base_size = 12) + theme(legend.position = "top")
```


# Model Fitting {.tabset}

We evaluate five classification models spanning a complexity spectrum: Naive Bayes (probabilistic baseline), Multinomial Logistic Regression (linear baseline), Random Forest and Gradient Boosting (ensemble methods), and a Deep Neural Network (DNN). All models use the same shared train/test split and are evaluated with identical metrics. Each model employs a tailored strategy for class imbalance. Model training and tuning were pre-computed via `run_all_models.R`; the final-model training code is shown in foldable blocks below and results are loaded from `.rds` files.

```{r load-results, eval=TRUE}
# Load all pre-computed results
gbm_res <- readRDS("results/gbm_3class.rds")
dnn_res <- readRDS("results/dnn_3class.rds")
nb_res  <- readRDS("results/nb_3class.rds")
rf_res  <- readRDS("results/rf_3class.rds")
lr_res  <- readRDS("results/lr_3class.rds")
comparison <- readRDS("results/model_comparison.rds")

# Shared confusion matrix plot function
plot_cm <- function(cm_obj, title) {
  cm_df <- as.data.frame(cm_obj$table)
  names(cm_df) <- c("Predicted", "Actual", "Freq")
  cm_df <- cm_df %>% group_by(Actual) %>% mutate(Pct = Freq / sum(Freq)) %>% ungroup()
  ggplot(cm_df, aes(x = Actual, y = Predicted, fill = Pct)) +
    geom_tile(color = "white", linewidth = 0.8) +
    geom_text(aes(label = sprintf("%d\n(%.0f%%)", Freq, Pct * 100)), size = 3.5, fontface = "bold") +
    scale_fill_gradient(low = "#E3F2FD", high = "#1565C0", labels = scales::percent) +
    labs(title = title, x = "Actual", y = "Predicted", fill = "Rate") +
    theme_minimal(base_size = 11)
}
```

## Gradient Boosting (GBM)

GBM fits an additive ensemble of shallow decision trees, with each tree correcting the residual errors of the previous ensemble. Class imbalance is handled via **sqrt-based case weights**. The final model uses the best hyperparameters selected by 3-fold CV (see Model Improvement for tuning details).

```{r gbm-final-code, eval=FALSE}
# ============================================================
# GBM: Final model with best hyperparameters (pre-computed)
# ============================================================
make_class_weights <- function(y) {
  y <- factor(y); cc <- table(y)
  raw_w <- sqrt(max(cc) / cc)
  as.numeric(raw_w[as.character(y)])
}

# Best params selected from 3-fold CV grid search (24 combos)
best_grid_gbm <- gbm_tuned_3$bestTune

# Train final model on full training set (no CV, just fit)
set.seed(123)
gbm_final_3 <- train(
  JobSat_Class ~ ., data = train_3,
  method = "gbm",
  trControl = trainControl(method = "none", classProbs = FALSE),
  tuneGrid = best_grid_gbm,
  weights = make_class_weights(train_3$JobSat_Class),
  verbose = FALSE
)

# Predict on held-out test set
gbm_pred_3 <- predict(gbm_final_3, newdata = test_3)
```

**Best parameters:** n.trees = `r gbm_res$best_params$n.trees`, depth = `r gbm_res$best_params$interaction.depth`, shrinkage = `r gbm_res$best_params$shrinkage`, min.node = `r gbm_res$best_params$n.minobsinnode`. **Imbalance strategy:** sqrt case weights.

```{r gbm-cm, eval=TRUE, fig.height=3.5}
plot_cm(gbm_res$confusion_matrix, "GBM — Confusion Matrix (Test Set)")
```

## Deep Neural Network (DNN)

A 3-layer feedforward network (Linear → ReLU → Dropout per layer) trained with **Focal Loss** ($FL = -\alpha_t(1-p_t)^\gamma \log p_t$, $\gamma = 2.0$), which dynamically down-weights easy samples and focuses on hard examples. The Neutral class receives an additional 1.5× weight boost. DNN-specific preprocessing includes one-hot encoding, Z-score standardisation (parameters from training set only), and value clipping to [-5, 5].

```{r dnn-final-code, eval=FALSE}
# ============================================================
# DNN: Final model with best hyperparameters (pre-computed)
# Requires: library(torch); library(luz)
# ============================================================

# DNN-specific preprocessing: one-hot + Z-score (train params only) + clip [-5, 5]
X_train_raw <- train_data %>% select(-JobSat_Class)
X_test_raw  <- test_data  %>% select(-JobSat_Class)
X_train_raw$.__split__ <- "train"; X_test_raw$.__split__ <- "test"
combined <- bind_rows(X_train_raw, X_test_raw) %>% mutate(across(where(is.character), as.factor))
split_col <- combined$.__split__; combined$.__split__ <- NULL
X_onehot <- model.matrix(~ . - 1, data = combined)
X_train_oh <- X_onehot[split_col == "train", ]; X_test_oh <- X_onehot[split_col == "test", ]
train_means <- apply(X_train_oh, 2, mean); train_sds <- apply(X_train_oh, 2, sd)
train_sds[train_sds == 0] <- 1
X_train_mat <- as.matrix(pmax(pmin(scale(X_train_oh, center=train_means, scale=train_sds), 5), -5))
X_test_mat  <- as.matrix(pmax(pmin(scale(X_test_oh, center=train_means, scale=train_sds), 5), -5))

# Class weights: sqrt weighting + Neutral x1.5 boost
dnn_counts <- table(factor(y_train_dnn, levels = 1:3))
dnn_weights <- as.numeric(sqrt(length(y_train_dnn) / (3 * dnn_counts)))
dnn_weights <- dnn_weights / sum(dnn_weights) * 3
dnn_weights[2] <- dnn_weights[2] * 1.5
dnn_weights <- dnn_weights / sum(dnn_weights) * 3

# Best params from 3-fold CV grid search (8 combos)
# Train final model: 80 epochs, lr scheduler, early stopping (patience=15)
set.seed(123); torch_manual_seed(123)
dnn_final_model <- dnn_model %>%
  setup(loss = focal_loss_module(weight = class_weights_tensor, gamma = 2.0),
        optimizer = optim_adam, metrics = list(luz_metric_accuracy())) %>%
  set_hparams(n_features = n_features, n_classes = 3,
              hidden1 = best_dims$h1, hidden2 = best_dims$h2,
              hidden3 = best_dims$h3, dropout_rate = best_dnn$dropout_rate) %>%
  set_opt_hparams(lr = best_dnn$lr, weight_decay = 1e-4) %>%
  fit(data = final_train_dl, epochs = 80, valid_data = final_test_dl,
      accelerator = accelerator(device_placement = TRUE, cpu = !cuda_is_available()),
      callbacks = list(
        luz_callback_lr_scheduler(torch::lr_reduce_on_plateau, patience=5, factor=0.5),
        luz_callback_early_stopping(patience = 15)), verbose = TRUE)
```

**Best parameters:** lr = `r dnn_res$best_params$lr`, dropout = `r dnn_res$best_params$dropout`, hidden = `r dnn_res$best_params$hidden`. **Imbalance strategy:** Focal Loss (gamma=2) + Neutral ×1.5 boost.

```{r dnn-cm, eval=TRUE, fig.height=3.5}
plot_cm(dnn_res$confusion_matrix, "DNN — Confusion Matrix (Test Set)")
```

## Naive Bayes (NB)

Naive Bayes estimates class-conditional feature probabilities under a conditional independence assumption. Since NB cannot accept case weights, class imbalance is addressed via **SMOTE oversampling** (over_ratio = 0.8) applied to the training set only. Test set remains unmodified.

```{r nb-final-code, eval=FALSE}
# ============================================================
# NB: Final model with best hyperparameters (pre-computed)
# Uses shared train_3/test_3 for fair comparison (not re-split)
# ============================================================
library(e1071); library(recipes); library(themis)

# Convert factors to numeric for SMOTE, keep target as factor
train_nb <- train_3; test_nb <- test_3
factor_cols_nb <- setdiff(names(train_nb)[sapply(train_nb, is.factor)], "JobSat_Class")
train_nb[factor_cols_nb] <- lapply(train_nb[factor_cols_nb], as.numeric)
test_nb[factor_cols_nb]  <- lapply(test_nb[factor_cols_nb], as.numeric)

# SMOTE oversampling (over_ratio = 0.8)
rec_nb <- recipe(JobSat_Class ~ ., data = train_nb) %>%
  step_smote(JobSat_Class, over_ratio = 0.8, seed = 123)
train_bal_nb <- bake(prep(rec_nb, training = train_nb), new_data = NULL)

# Best params from 3-fold CV (6 combos: laplace x usekernel)
set.seed(123)
nb_model_3 <- train(
  JobSat_Class ~ ., data = train_bal_nb, method = "naive_bayes",
  trControl = trainControl(method="cv", number=3, classProbs=TRUE,
                           summaryFunction=multiClassSummary),
  tuneGrid = expand.grid(laplace=c(0,1,5), usekernel=c(FALSE,TRUE), adjust=1),
  metric = "Mean_F1"
)

nb_pred_3 <- predict(nb_model_3, newdata = test_nb)
```

**Best parameters:** laplace = `r nb_res$best_params$laplace`, usekernel = `r nb_res$best_params$usekernel`. **Imbalance strategy:** SMOTE (over_ratio = 0.8).

```{r nb-cm, eval=TRUE, fig.height=3.5}
plot_cm(nb_res$confusion_matrix, "Naive Bayes — Confusion Matrix (Test Set)")
```

## Random Forest (RF)

Random Forest aggregates predictions from an ensemble of de-correlated decision trees. Two imbalance strategies are compared: **power-based class weights** (power = 0.75/0.85/0.95) and **downsampling**. The best combination is selected via 3-fold CV Macro F1. The `extratrees` split rule provides extra randomisation.

```{r rf-final-code, eval=FALSE}
# ============================================================
# RF: Best model selected from multiple strategies (pre-computed)
# ============================================================
library(ranger)

make_class_weights_rf <- function(y, weight_power = 0.75) {
  cc <- table(y); as.numeric((max(cc) / cc) ^ weight_power)
}

# Grid: mtry={6,8,10,12,16} x min.node.size={10,15,20} x splitrule=extratrees
# Compared: 3 weight powers (0.75, 0.85, 0.95) + downsample (~60 total combos)
# All evaluated via 3-fold CV Macro F1, num.trees = 100

for (wp in c(0.75, 0.85, 0.95)) {
  rf_fit <- train(
    JobSat_Class ~ ., data = train_3, method = "ranger",
    trControl = ctrl_rf, tuneGrid = rf_grid_w, metric = "Macro_F1",
    num.trees = 100, importance = "impurity",
    respect.unordered.factors = "order",
    class.weights = make_class_weights_rf(train_3$JobSat_Class, wp))
}

rf_down <- train(
  JobSat_Class ~ ., data = train_3, method = "ranger",
  trControl = trainControl(method="cv", number=3, sampling="down"),
  tuneGrid = rf_grid_d, metric = "Macro_F1", num.trees = 100)

# Select best across all strategies
best_rf <- all_rf_models[[best_rf_id]]
rf_pred_3 <- predict(best_rf, newdata = test_3)
```

**Best parameters:** mtry = `r rf_res$best_params$mtry`, min.node.size = `r rf_res$best_params$min.node.size`, strategy = `r rf_res$best_params$model_id`. **Imbalance strategy:** `r rf_res$best_params$model_id`.

```{r rf-cm, eval=TRUE, fig.height=3.5}
plot_cm(rf_res$confusion_matrix, "Random Forest — Confusion Matrix (Test Set)")
```

## Multinomial Logistic Regression (LR)

Multinomial logistic regression extends binary logistic regression to multiple classes via a softmax link. Features are dummy-encoded (`fullRank = TRUE`) and standardised (center + scale, parameters from training set only). Class imbalance is handled via **power-weighted case weights** (power = 0.75).

```{r lr-final-code, eval=FALSE}
# ============================================================
# LR: Final model with best hyperparameters (pre-computed)
# ============================================================
library(nnet)

make_case_weights_lr <- function(y, weight_power = 0.75) {
  cc <- table(y); cw <- (max(cc) / cc) ^ weight_power
  as.numeric(cw[as.character(y)])
}

# Dummy encoding + NZV removal on dummy columns
dummy_lr <- dummyVars(~ ., data = x_train_lr, fullRank = TRUE)
x_train_dummy <- predict(dummy_lr, newdata = x_train_lr) %>% as.data.frame()
x_test_dummy  <- predict(dummy_lr, newdata = x_test_lr) %>% as.data.frame()
nzv_idx <- nearZeroVar(x_train_dummy, saveMetrics = FALSE)
if (length(nzv_idx) > 0) {
  x_train_dummy <- x_train_dummy[, -nzv_idx]; x_test_dummy <- x_test_dummy[, -nzv_idx]
}

# Grid search: decay = {0, 0.001, 0.01, 0.1}, 3-fold CV
set.seed(123)
lr_model_3 <- train(
  x = x_train_dummy, y = y_train_lr, method = "multinom",
  trControl = trainControl(method="cv", number=3, summaryFunction=macro_f1_summary, classProbs=FALSE),
  tuneGrid = expand.grid(decay = c(0, 0.001, 0.01, 0.1)),
  metric = "Macro_F1", preProcess = c("center","scale"),
  weights = make_case_weights_lr(y_train_lr, 0.75),
  trace = FALSE, maxit = 300, MaxNWts = 100000)

lr_pred_3 <- predict(lr_model_3, newdata = x_test_dummy)
```

**Best parameters:** decay = `r lr_res$best_params$decay`. **Imbalance strategy:** power case weights (0.75).

```{r lr-cm, eval=TRUE, fig.height=3.5}
plot_cm(lr_res$confusion_matrix, "Logistic Regression — Confusion Matrix (Test Set)")
```


# Model Improvement {.tabset}

## Feature Selection and Engineering

Starting from 172 raw survey columns, we applied a multi-stage feature selection pipeline. All decisions were based exclusively on training data.

```{r feature-funnel, eval=TRUE}
funnel_df <- data.frame(
  Stage = c("Raw survey columns",
            "Domain-relevant candidate selection",
            "After high-missingness removal (>80%)",
            "After multi-response binarisation + LeavingScore",
            "After near-zero variance removal",
            "After high-correlation removal (r > 0.90)",
            "Final feature set used by all models"),
  Features = c(172, 22, "~20", "~50+ (binary expansion)", "varies", "varies",
               ncol(train_3) - 1))
kable(funnel_df, col.names = c("Stage", "Feature Count"), align = "lr") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = TRUE, font_size = 12)
```

**LeavingScore — Engineered Ordinal Feature.** The categorical `NewRole` variable captures whether a developer has considered or undergone a career change. We mapped its 5 response levels to an ordinal 0–4 scale: 0 = voluntarily transitioned (already acted, satisfaction rebounds), 1 = neither considered nor transitioned (most stable), 2 = involuntarily transitioned, 3 = somewhat considered changing, 4 = strongly considered changing (most dissatisfied). The rationale is that one-hot encoding of NewRole loses the inherent ordering of career-change propensity, while a single numeric column preserves the monotonic signal and benefits all model types — linear models can capture the effect with one coefficient, while tree-based models gain an efficient split variable.

```{r leaving-score-plot, eval=TRUE, fig.height=3}
ls_by_class <- train_3 %>%
  group_by(JobSat_Class) %>%
  summarise(Mean_LS = round(mean(LeavingScore, na.rm = TRUE), 2), .groups = "drop")

ggplot(ls_by_class, aes(x = JobSat_Class, y = Mean_LS, fill = JobSat_Class)) +
  geom_col(width = 0.5) +
  geom_text(aes(label = Mean_LS), vjust = -0.3, size = 4, fontface = "bold") +
  scale_fill_manual(values = c(Dissatisfied="#E53935", Neutral="#FB8C00", Satisfied="#43A047")) +
  labs(title = "LeavingScore Mean by Satisfaction Class",
       subtitle = "Perfect monotonic decrease validates the ordinal encoding design",
       x = NULL, y = "Mean LeavingScore (0-4)") +
  theme_minimal(base_size = 12) + theme(legend.position = "none") + ylim(0, 3.5)
```

**3-Class vs 5-Class.** We chose 3-class over 5-class because adjacent 5-class boundaries (e.g., JobSat 3 vs 4, or 6 vs 7) represent only a 1-point difference on a subjective scale. The 3-class formulation provides clearer separation (0–3 vs 4–6 vs 7–10) and reduces the extreme minority class problem (Very_Dissatisfied had only ~326 samples in 5-class).

## Hyperparameter Tuning

All five models were tuned using **3-fold cross-validation** with **Macro F1** as the primary selection metric. Complete tuning code was executed in advance via `run_all_models.R`; the key tuning logic is shown in foldable blocks below.

```{r gbm-tuning-code, eval=FALSE}
# ============================================================
# GBM tuning: 24 combos on stratified subset, 3-fold CV
# ============================================================
gbm_grid_3 <- expand.grid(
  n.trees           = c(500, 800, 1000),
  interaction.depth = c(3, 5),
  shrinkage         = c(0.03, 0.05),
  n.minobsinnode    = c(10, 20)
)

# Use stratified subset for efficiency (up to 12000 rows)
set.seed(123)
tune_p <- min(1, 12000 / nrow(train_3))
tune_idx <- createDataPartition(train_3$JobSat_Class, p = tune_p, list = FALSE)
tune_data <- train_3[tune_idx, ]

gbm_tuned_3 <- train(
  JobSat_Class ~ ., data = tune_data, method = "gbm",
  trControl = trainControl(method = "cv", number = 3,
    summaryFunction = macro_f1_summary, savePredictions = "final"),
  tuneGrid = gbm_grid_3, metric = "Macro_F1",
  weights = make_class_weights(tune_data$JobSat_Class), verbose = FALSE)
```

```{r dnn-tuning-code, eval=FALSE}
# ============================================================
# DNN tuning: 8 combos x 3 folds = 24 training runs
# Manual CV loop (torch does not integrate with caret)
# ============================================================
dnn_grid <- expand.grid(
  lr           = c(0.001, 0.0005),
  dropout_rate = c(0.3, 0.5),
  hidden_size  = c("large", "small"),  # large = 128/64/32, small = 64/32/16
  stringsAsFactors = FALSE
)

fold_index_dnn <- createFolds(factor(y_train_dnn, levels = 1:3), k = 3, list = TRUE)

for (i in seq_len(nrow(dnn_grid))) {
  for (k in 1:3) {
    # Per-fold: split data, compute fold-specific class weights (sqrt + Neutral x1.5)
    # Create DataLoader (batch_size = 256)
    # Train with Focal Loss (gamma=2.0), Adam (weight_decay=1e-4)
    # Max 80 epochs, early stopping patience = 12
    # Evaluate Macro F1 on validation fold
  }
  # Record mean +/- SD of CV Macro F1
}
# Select combo with highest mean CV Macro F1
```

```{r nb-tuning-code, eval=FALSE}
# ============================================================
# NB tuning: 6 combos, 3-fold CV on SMOTE-balanced training set
# ============================================================
tune_grid_nb <- expand.grid(
  laplace = c(0, 1, 5), usekernel = c(FALSE, TRUE), adjust = 1)

nb_model_3 <- train(
  JobSat_Class ~ ., data = train_bal_nb, method = "naive_bayes",
  trControl = trainControl(method="cv", number=3, classProbs=TRUE,
                           summaryFunction=multiClassSummary),
  tuneGrid = tune_grid_nb, metric = "Mean_F1")
```

```{r rf-tuning-code, eval=FALSE}
# ============================================================
# RF tuning: ~60 combos across 4 strategies, 3-fold CV
# ============================================================
rf_grid_w <- expand.grid(
  mtry = unique(pmin(ncol(train_3)-1, c(6,8,10,12,16))),
  splitrule = "extratrees", min.node.size = c(10, 15, 20))

# Class-weighted RF with 3 weight powers
for (wp in c(0.75, 0.85, 0.95)) {
  train(JobSat_Class ~ ., data = train_3, method = "ranger",
    trControl = ctrl_rf, tuneGrid = rf_grid_w, metric = "Macro_F1",
    num.trees = 100, class.weights = make_class_weights_rf(train_3$JobSat_Class, wp))
}

# Downsampled RF for comparison
train(JobSat_Class ~ ., data = train_3, method = "ranger",
  trControl = trainControl(method="cv", number=3, sampling="down"),
  tuneGrid = rf_grid_d, metric = "Macro_F1", num.trees = 100)
```

```{r lr-tuning-code, eval=FALSE}
# ============================================================
# LR tuning: 4 decay values, 3-fold CV
# ============================================================
train(x = x_train_dummy, y = y_train_lr, method = "multinom",
  trControl = trainControl(method="cv", number=3, summaryFunction=macro_f1_summary),
  tuneGrid = expand.grid(decay = c(0, 0.001, 0.01, 0.1)),
  metric = "Macro_F1", preProcess = c("center","scale"),
  weights = make_case_weights_lr(y_train_lr, 0.75),
  trace = FALSE, maxit = 300, MaxNWts = 100000)
```

The table below summarises all tuning configurations and results:

```{r tuning-summary, eval=TRUE}
tuning_summary <- data.frame(
  Model = c("GBM", "DNN", "Naive Bayes", "Random Forest", "Logistic Regression"),
  Search_Space = c(
    "n.trees={500,800,1000} x depth={3,5} x shrinkage={0.03,0.05} x min.node={10,20}",
    "lr={0.001,0.0005} x dropout={0.3,0.5} x hidden={large(128/64/32),small(64/32/16)}",
    "laplace={0,1,5} x usekernel={TRUE,FALSE}",
    "mtry={6,8,10,12,16} x min.node={10,15,20} x weight_power={0.75,0.85,0.95} + downsample",
    "decay={0,0.001,0.01,0.1}"),
  Combos = c(24, 8, 6, "~60", 4),
  CV = rep("3-fold", 5),
  Imbalance = c("Sqrt case weights", "Focal Loss (gamma=2) + Neutral x1.5",
                "SMOTE (ratio=0.8)", "Power class weights + downsample",
                "Power case weights (0.75)"),
  Best_Params = c(
    paste0("trees=", gbm_res$best_params$n.trees, " depth=", gbm_res$best_params$interaction.depth,
           " lr=", gbm_res$best_params$shrinkage, " node=", gbm_res$best_params$n.minobsinnode),
    paste0("lr=", dnn_res$best_params$lr, " drop=", dnn_res$best_params$dropout,
           " net=", dnn_res$best_params$hidden),
    paste0("laplace=", nb_res$best_params$laplace, " kernel=", nb_res$best_params$usekernel),
    paste0("mtry=", rf_res$best_params$mtry, " node=", rf_res$best_params$min.node.size,
           " ", rf_res$best_params$model_id),
    paste0("decay=", lr_res$best_params$decay)),
  Test_MacroF1 = c(gbm_res$overall_metrics$Macro_F1, dnn_res$overall_metrics$Macro_F1,
                   nb_res$overall_metrics$Macro_F1, rf_res$overall_metrics$Macro_F1,
                   lr_res$overall_metrics$Macro_F1))

kable(tuning_summary,
  col.names = c("Model","Search Space","Combos","CV","Imbalance Strategy","Best Parameters","Test Macro F1"),
  align = "llcclcr") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = TRUE, font_size = 11) %>%
  column_spec(7, bold = TRUE)
```

**Key findings:** GBM achieved the highest test Macro F1 (`r gbm_res$overall_metrics$Macro_F1`), benefiting from iterative error correction and native handling of mixed feature types. DNN performed comparably (`r dnn_res$overall_metrics$Macro_F1`) — Focal Loss effectively addressed the class imbalance by focusing on hard-to-classify samples. Naive Bayes (`r nb_res$overall_metrics$Macro_F1`) served as a useful lower bound, confirming that the conditional independence assumption is severely violated in this dataset and that more complex models are justified.

<small>*The complete model training and tuning pipeline (`run_all_models.R`, 1239 lines) is included in the foldable code block below for full reproducibility.*</small>

```{r run-all-models-full, eval=FALSE, code=readLines("run_all_models.R")}
```


# Model Evaluation & Selection {.tabset}

## Overall Comparison

```{r comparison-table, eval=TRUE}
kable(comparison,
  col.names = c("Model","Accuracy","Kappa","Macro F1","Weighted F1","Macro Precision","Macro Recall"),
  align = "lrrrrrr", digits = 4) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = TRUE) %>%
  row_spec(1, bold = TRUE, background = "#E3F2FD")
```

```{r macro-f1-bar, eval=TRUE, fig.height=3.5}
ggplot(comparison, aes(x = reorder(Model, Macro_F1), y = Macro_F1, fill = Model)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = sprintf("%.4f", Macro_F1)), hjust = -0.1, size = 3.5, fontface = "bold") +
  coord_flip() +
  scale_fill_manual(values = c("GBM"="#1565C0", "DNN"="#E65100",
    "Multinomial Logistic Regression"="#2E7D32", "Random Forest"="#6A1B9A", "Naive Bayes"="#C62828")) +
  labs(title = "Test Set Macro F1 Comparison (Primary Metric)", x = NULL, y = "Macro F1") +
  theme_minimal(base_size = 12) + theme(legend.position = "none") +
  ylim(0, max(comparison$Macro_F1) * 1.15)
```

```{r per-class-f1, eval=TRUE, fig.height=4}
all_class <- bind_rows(
  gbm_res$class_metrics %>% mutate(Model = "GBM"),
  dnn_res$class_metrics %>% mutate(Model = "DNN"),
  nb_res$class_metrics  %>% mutate(Model = "Naive Bayes"),
  rf_res$class_metrics  %>% mutate(Model = "Random Forest"),
  lr_res$class_metrics  %>% mutate(Model = "Logistic Regression"))

ggplot(all_class, aes(x = Class, y = F1, fill = Model)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(aes(label = sprintf("%.2f", F1)),
            position = position_dodge(0.7), vjust = -0.3, size = 2.8) +
  scale_fill_manual(values = c("GBM"="#1565C0", "DNN"="#E65100",
    "Logistic Regression"="#2E7D32", "Random Forest"="#6A1B9A", "Naive Bayes"="#C62828")) +
  labs(title = "Per-Class F1 Score by Model",
       subtitle = "Dissatisfied = hardest (minority class); Satisfied = easiest (majority class)",
       x = NULL, y = "F1 Score") +
  theme_minimal(base_size = 12) + theme(legend.position = "top") + ylim(0, 1)
```

## Final Model Selection

Based on the evaluation, **Gradient Boosting (GBM)** is selected as the final model:

1. **Highest Macro F1** (`r gbm_res$overall_metrics$Macro_F1`): best balance across all three classes, which is critical given the class imbalance.
2. **Highest Kappa** (`r gbm_res$overall_metrics$Kappa`): strongest agreement beyond chance among all five models.
3. **Balanced per-class performance**: achieves reasonable F1 for all classes, including the challenging Dissatisfied minority.
4. **Interpretability**: variable importance is readily available, providing actionable domain insights.

```{r gbm-varimp, eval=TRUE, fig.height=4.5}
plot(gbm_res$variable_importance, top = 15,
     main = "GBM — Top 15 Most Important Predictors")
```

The variable importance plot reveals that **LeavingScore** (career-change propensity) is a dominant predictor, followed by Country, AI-related attitudes, and compensation. This aligns with domain knowledge: developers actively considering leaving their role are strongly associated with lower satisfaction.

**Discussion.** All five models achieve Macro F1 in the 0.44–0.47 range, reflecting the inherent difficulty of predicting subjective job satisfaction from survey features. The relatively narrow performance gap between the simplest useful model (Logistic Regression, `r lr_res$overall_metrics$Macro_F1`) and the best (GBM, `r gbm_res$overall_metrics$Macro_F1`) suggests that the signal-to-noise ratio in the dataset is the primary bottleneck rather than model capacity. Nevertheless, the structured comparison demonstrates that ensemble methods (GBM, RF) consistently outperform linear and probabilistic baselines, and that thoughtful imbalance handling — whether via class weights, Focal Loss, or SMOTE — is essential for achieving non-trivial Macro F1 in skewed multi-class settings.
